---
title: "BIOSTAT 203B HW1"
author: "Benson Wu"
date: "1/14/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Display machine information for reproducibility:
```{r, eval=T}
sessionInfo()
```

## Q1. Git/GitHub

**No handwritten homework reports are accepted for this course.** We work with Git and GitHub. Efficient and abundant use of Git, e.g., frequent and well-documented commits, is an important criterion for grading your homework.

1. Apply for the [Student Developer Pack](https://education.github.com/pack) at GitHub using your UCLA email.

**Solution**: Done. 

2. Create a **private** repository `biostat-203b-2021-winter` and add `Hua-Zhou`, `Chris-German` and `ElvisCuiHan` as your collaborators with write permission.

**Solution**: Done. 

3. Top directories of the repository should be `hw1`, `hw2`, ... Maintain two branches `master` and `develop`. The `develop` branch will be your main playground, the place where you develop solution (code) to homework problems and write up report. The `master` branch will be your presentation area. Submit your homework files (R markdown file `Rmd`, `html` file converted from R markdown, all code and data sets to reproduce results) in `master` branch.

**Solution**: Done. 

4. After each homework due date, teaching assistant and instructor will check out your master branch for grading. Tag each of your homework submissions with tag names `hw1`, `hw2`, ... Tagging time will be used as your submission time. That means if you tag your `hw1` submission after deadline, penalty points will be deducted for late submission.

**Solution**: Done. 

5. After this course, you can make this repository public and use it to demonstrate your skill sets on job market.

**Solution**: Sounds good. 

## Q2. Linux Shell Commands

1. This exercise (and later in this course) uses the [MIMIC-IV data](https://mimic-iv.mit.edu), a freely accessible critical care database developed by the MIT Lab for Computational Physiology. Follow the instructions at <https://mimic-iv.mit.edu/docs/access/> to (1) complete the CITI `Data or Specimens Only Research` course and (2) obtain the PhysioNet credential for using the MIMIC-IV data. Display the verification links to your completion report and completion certificate here. (Hint: The CITI training takes a couple hours and the PhysioNet credentialing takes a couple days; do not leave it to the last minute.)

**Solution**: Completed and received PhysioNet credential to use MIMIC-IV data. Here is my [course completion report](https://www.citiprogram.org/verify/?k7d62b2a6-a528-4c9f-b08e-3b27874318b8-40352501) and [course completion certificate](https://www.citiprogram.org/verify/?w57643131-f682-4a30-bff8-0794fe0eacf7-40352501)

2. The `/usr/203b-data/mimic-iv/` folder on teaching server contains data sets from MIMIC-IV. Refer to <https://mimic-iv.mit.edu/docs/datasets/> for details of data files.  
    ```{bash}
    ls -l /usr/203b-data/mimic-iv
    ```
Please, do **not** put these data files into Git; they are big. Do **not** copy them into your directory. Do **not** decompress the gz data files. These create unnecessary big files on storage and are not big data friendly practices. Just read from the data folder `/usr/203b-data/mimic-iv` directly in following exercises. 

**Solution**: Okay, will do.

    Use Bash commands to answer following questions.

3. Display the contents in the folders `core`, `hosp`, `icu`. What are the functionalities of the bash commands `zcat`, `zless`, `zmore`, and `zgrep`? 

**Solutions** 
```{bash}
ls -al /usr/203b-data/mimic-iv/core
```

```{bash}
ls -al /usr/203b-data/mimic-iv/hosp
```

```{bash}
ls -al /usr/203b-data/mimic-iv/icu
```

`zcat`: Displays the contents of a compressed files
`zless`: `zmore`, and 
`zgrep`: Search the compressed file

**Solution**

4. What's the output of following bash script?
    ```{bash, eval=F}
    for datafile in /usr/203b-data/mimic-iv/core/*.gz
      do
        ls -l $datafile
      done
    ```
Display the number of lines in each data file using a similar loop.

4. Display the first few lines of `admissions.csv.gz`. How many rows are in this data file? How many unique patients (identified by `subject_id`) are in this data file? What are the possible values taken by each of the variable `admission_type`, `admission_location`, `insurance`, `language`, `marital_status`, and `ethnicity`? Also report the count for each unique value of these variables. (Hint: combine Linux commands `zcat`, `head`/`tail`, `awk`, `uniq`, `wc`, and so on.)

## Q3. Who's popular in Price and Prejudice

1. You and your friend just have finished reading *Pride and Prejudice* by Jane Austen. Among the four main characters in the book, Elizabeth, Jane, Lydia, and Darcy, your friend thinks that Darcy was the most mentioned. You, however, are certain it was Elizabeth. Obtain the full text of the novel from <http://www.gutenberg.org/cache/epub/42671/pg42671.txt> and save to your local folder. 
    ```{bash, eval=FALSE}
    curl http://www.gutenberg.org/cache/epub/42671/pg42671.txt > pride_and_prejudice.txt
    ```
Do **not** put this text file `pride_and_prejudice.txt` in Git. Using a `for` loop, how would you tabulate the number of times each of the four characters is mentioned?

2. What's the difference between the following two commands?
    ```{bash eval=FALSE}
    echo 'hello, world' > test1.txt
    ```
    and
    ```{bash eval=FALSE}
    echo 'hello, world' >> test2.txt
    ```

3. Using your favorite text editor (e.g., `vi`), type the following and save the file as `middle.sh`:
    ```{bash eval=FALSE}
    #!/bin/sh
    # Select lines from the middle of a file.
    # Usage: bash middle.sh filename end_line num_lines
    head -n "$2" "$1" | tail -n "$3"
    ```
Using `chmod` make the file executable by the owner, and run 
    ```{bash eval=FALSE}
    ./middle.sh pride_and_prejudice.txt 20 5
    ```
Explain the output. Explain the meaning of `"$1"`, `"$2"`, and `"$3"` in this shell script. Why do we need the first line of the shell script?

### Q4. More fun with Linux

Try these commands in Bash and interpret the results: `cal`, `cal 2021`, `cal 9 1752` (anything unusual?), `date`, `hostname`, `arch`, `uname -a`, `uptime`, `who am i`, `who`, `w`, `id`, `last | head`, `echo {con,pre}{sent,fer}{s,ed}`, `time sleep 5`, `history | tail`.

**Solution**
```{bash}
cal
```
`cal` shows the calendar of the current month with the current date highlighted

```{bash}
cal 2021
```

`cal 2021`: shows the calendar of 2021 year with the current date highlighted


```{bash}
cal 9 1752
```

`cal 9 1752`: shows the calendar for September 1752, but the 3rd-13th are missing from the calendar!

```{bash}
date
```

`date`: gives the current day of the week, date, coordinated universal time, and year

```{bash}
hostname
```

`hostname`: gives the domain name system. In this case it is "ucla-biostat-203b"

```{bash}
arch
```
`arch`: this prints the architecure of the machine. 

```{bash}
uname -a
```

`uname -a`: this prints informartion about the operating system and machine hardware. 

```{bash}
uptime
```

`uptime`: This gives the current time, how many users are on the server, and load average in the last 1, 5, and 15 minutes 

```{bash}
who am i
```

`who am i`: This returns nothing, but the following works

```{bash}
whoami
```

`whoami`: returns username of curent user

```{bash}
who
```

`who`: returns the usernames of users who are on the server, their terminal, log-in time, and what host they're using the server from.

```{bash}
w
```

`w`: Header contains same information that `uptime` returns. Below the header contains the usernames of users who are on the server, their terminal type, host name, login time, idle time, JCPU time, PCPU time, and what type of command line they are using. 

```{bash}
id
```

`id`: Returns user ID, group ID, supplemental group ID, and the security context (https://linuxize.com/post/id-command-in-linux/)

```{bash}
last | head
```

`last | head`: show the head (first 10 entries by default) of the usernames of the users who logged in to the server, the terminal type, IP address, the date and time of log in, and the duration of the time that they were logged in. 

```{bash}
echo {con,pre}{sent,fer}{s,ed}
```


`echo {con,pre}{sent,fer}{s,ed}`: Displays the combination of the three sets of letters (maintaining the order of the sets though)

```{bash}
time sleep 5
```
`time sleep 5`: gives the amount of time that it takes to run `sleep 5`, which says to delay execution of a script for 5 seconds, so expectedly it the real time displayed is ~5 seconds. 

```{bash}
history | tail
```

`history | tail`: Gives you the tail (last 10 by default) commands that you ran 
